---
title: "SupFig_2456"
output: html_document
date: "2024-02-14"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(LDM)
library(vegan)
library(dplyr)
library(tidyr)
library(ggplot2)
```

Load metadata
```{r}
meta<-read.csv("../data/first_batch/metadata.csv")
```

Generate underlying LDM for SF2 / M v F baseline
```{r}
# Select proper samples and add group name
met<- meta %>% filter(Time == "1")
strains<-read.csv("../data/first_batch/Figure3/Male_vs_female_baseline_STRAINS.csv")

# Sum strains within species
strains2 <- strains %>%
  group_by(Species) %>%
  summarise(across(.cols = 9:29, .fns = sum, na.rm = TRUE))

# Take set for graphing
bugs<-strains2[2:21]
tax<-strains2$Species

# Fix colnames
colnames(bugs)<-gsub("_.*", "", colnames(bugs))
colnames(bugs)<-gsub("X", "",  colnames(bugs))

# Make sure column order matches metadata 
column_order <- met$mouseID
bugs <- bugs[, order(match(colnames(bugs), column_order))]
identical(colnames(bugs), as.character(met$mouseID))

# Strip NA before LDM
bugs[is.na(bugs)] <- 0

fit1<-ldm(formula=bugs ~ Sex, data=met, dist.method = "bray", seed=12345)
fit1$p.global.omni
fit1$detected.otu.omni # The OTUs significantly associated with each covariate

#Make summary table for SEX
w1 = match(fit1$detected.otu.omni[[1]], names(fit1$q.otu.omni))
o = w1[order(fit1$p.otu.omni[w1])]
summary.tab = data.frame(raw.pvalue=signif(fit1$p.otu.omni[o],3),
                       adj.pvalue=signif(fit1$q.otu.omni[o],3),
                       mean.freq=signif(fit1$mean.freq[o],3),
                       direction=t(ifelse(fit1$beta>0, "+", "-"))[o],
                       otu.name=names(fit1$q.otu.omni)[o], # can use taxonomy assignment
                       row.names=NULL)
colnames(summary.tab)[4] = paste("direction.", rownames(fit1$beta), sep="")

# fix taxonomy
tax_index<-data.frame(tax)
tax_index<- cbind(row_number = 1:nrow(tax_index), tax_index)
tax_index$row_number = as.character(tax_index$row_number)
colnames(tax_index)[2] = "Species"
summary.tab2<-left_join(summary.tab, tax_index, by = c("otu.name"="row_number"))

write.csv(summary.tab2, "../results/Sex-Time1-associated-ldm.csv")
```

Generate SupFig 2 from LDM outputs while we still have proper strains2 and metadata file
```{r}
# Load LDM results for sex
data<-read.csv("../results/Sex-Time1-associated-ldm.csv")
data <- arrange(data, desc(mean.freq))
data$p.adj<-ifelse(data$adj.pvalue < 0.01, "<0.01", ifelse(data$adj.pvalue < 0.05, "<0.05", "0.05 - 0.1"))

# Plot only significant species from LDM analysis
toplot<-strains2 %>% filter(Species %in% data$Species)
toplot$Taxonomy.Id = NULL

colnames(toplot)<-gsub("_.*", "", colnames(toplot))
colnames(toplot)<-gsub("X", "",  colnames(toplot))

# Reshape data
data_long <- pivot_longer(toplot, cols = -c(Species), names_to = "subject", values_to = "value")
# add metadata back for groupings

met$mouseID<-as.character(met$mouseID)
colnames(data_long)[2] = "mouseID"
df3 <- data_long %>% left_join(met, by = "mouseID") %>% select(mouseID, Sex, Species, value)

# Calculate freqs & SE
freqs <- df3 %>% group_by(Species, Sex) %>%
  summarize(mean=mean(value), sd=sd(value))
freqs$se = freqs$sd / sqrt(10)
freqs<-left_join(freqs, data)
freqs <- arrange(freqs, desc(mean.freq))
freqs$sig<-ifelse(freqs$adj.pvalue < 0.001, "***", ifelse(freqs$adj.pvalue < 0.01, "**", (ifelse(freqs$adj.pvalue < 0.05, "*", "-"))))

# Make plot
p<-ggplot(freqs,  aes(x = reorder(Species, mean.freq), y     = mean, color = Sex)) + geom_point() + geom_errorbar(aes(ymin  = mean - se, ymax  = mean + se))  + coord_flip()  + scale_y_sqrt() + xlab("Species") + ylab("Mean Abundance") +   scale_colour_manual(values=c("red", "blue"))

p + geom_text(aes(y = 0.23, x = freqs$Species), label = freqs$sig, color="black")

ggsave("../results/SupFig2.pdf", width=12, height=8)
```

Generate underlying LDM for SF6 / Fem T vs Veh wk 4
```{r}
# Select proper samples and add group name
met<- meta %>% filter(Time == "4") %>% filter(Sex=="Female")
strains<-read.csv("../data/first_batch/Figure3/Strains GAHT 4 wks Females only.csv")
met$Group<-ifelse(met$trt == "Test", "T", "Veh")
met$Group = factor(met$Group, levels = c("Veh", "T"))

 strains2 <- strains %>%
  group_by(Species) %>%
  summarise(across(.cols = 9:28, .fns = sum, na.rm = TRUE))

# Take set for graphing
bugs<-strains2[2:21]
tax<-strains2$Species

colnames(bugs)<-gsub("_.*", "", colnames(bugs))
colnames(bugs)<-gsub("X", "",  colnames(bugs))

# fix order of cols
column_order <- met$mouseID
bugs <- bugs[, order(match(colnames(bugs), column_order))]

# Check order
identical(as.character(met$mouseID), colnames(bugs))

# Strip NA before LDM
bugs[is.na(bugs)] <- 0

fit1<-ldm(formula=bugs ~ Group, data=met, dist.method = "bray", seed=1)
fit1$p.global.omni
fit1$detected.otu.omni # The OTUs significantly associated with each covariate

#Make summary table for SEX
w1 = match(fit1$detected.otu.omni[[1]], names(fit1$q.otu.omni))
o = w1[order(fit1$p.otu.omni[w1])]
summary.tab = data.frame(raw.pvalue=signif(fit1$p.otu.omni[o],3),
                       adj.pvalue=signif(fit1$q.otu.omni[o],3),
                       mean.freq=signif(fit1$mean.freq[o],3),
                       direction=t(ifelse(fit1$beta>0, "+", "-"))[o],
                       otu.name=names(fit1$q.otu.omni)[o], # can use taxonomy assignment
                       row.names=NULL)
colnames(summary.tab)[4] = paste("direction.", rownames(fit1$beta), sep="")

# fix taxonomy
tax_index<-data.frame(tax)
tax_index<- cbind(row_number = 1:nrow(tax_index), tax_index)
tax_index$row_number = as.character(tax_index$row_number)
colnames(tax_index)[2] = "Species"
summary.tab2<-left_join(summary.tab, tax_index, by = c("otu.name"="row_number"))

#write.csv(summary.tab2, "../results/FemGAHT-Time4-associated-ldm.csv")
```

Generate SupFig 6 from LDM outputs while we still have proper strains2 and metadata file
```{r}
# Load LDM results for sex
data<-read.csv("../results/FemGAHT-Time4-associated-ldm.csv")
data <- arrange(data, desc(mean.freq))
data$p.adj<-ifelse(data$adj.pvalue < 0.01, "<0.01", ifelse(data$adj.pvalue < 0.05, "<0.05", "0.05 - 0.1"))

# Plot only significant species from LDM analysis
toplot<-strains2 %>% filter(Species %in% data$Species)
toplot$Taxonomy.Id = NULL

colnames(toplot)<-gsub("_.*", "", colnames(toplot))
colnames(toplot)<-gsub("X", "",  colnames(toplot))

# Reshape data
data_long <- pivot_longer(toplot, cols = -c(Species), names_to = "subject", values_to = "value")
# add metadata back for groupings

met$mouseID<-as.character(met$mouseID)
colnames(data_long)[2] = "mouseID"
df3 <- data_long %>% left_join(met, by = "mouseID") %>% select(mouseID, Group, Species, value)

# Calculate freqs & SE
freqs <- df3 %>% group_by(Species, Group) %>%
  summarize(mean=mean(value), sd=sd(value))
freqs$se = freqs$sd / sqrt(10)
freqs<-left_join(freqs, data)
freqs <- arrange(freqs, desc(mean.freq))
freqs$sig<-ifelse(freqs$adj.pvalue < 0.001, "***", ifelse(freqs$adj.pvalue < 0.01, "**", (ifelse(freqs$adj.pvalue < 0.05, "*", "-"))))

# Make plot
p<-ggplot(freqs,  aes(x = reorder(Species, mean.freq), y     = mean, color = Group)) + geom_point() + geom_errorbar(aes(ymin  = mean - se, ymax  = mean + se))  + coord_flip()  + scale_y_sqrt() + xlab("Species") + ylab("Mean Abundance") +   scale_colour_manual(values=c("darkorange", "deeppink"))

p + geom_text(aes(y = 0.23, x = freqs$Species), label = freqs$sig, color="black")

ggsave("../results/SupFig6.pdf", width=12, height=8)
```

LDMs for male GAHT 4 wk
```{r}
met<- meta %>% filter(Time ==4) %>% filter(Sex == "Male")
met$Group<-ifelse(met$trt == "Estrogen", "Orx.E2", ifelse(met$orch =="Yes", "Orx.Veh", "Sham.Veh"))
met$Group = factor(met$Group, levels = c("Sham.Veh", "Orx.Veh", "Orx.E2"))
# Update dropbox link
strains<-read.csv("../data/first_batch/Figure3/Strains GAHT 4 wks males only.csv")
metacyc<-read.csv("../data/first_batch/Figure3/MetaCyc males 4 wks.csv")

mtax<-metacyc$Name
metacyc<-metacyc[2:30]
colnames(metacyc)<-gsub("_.*", "", colnames(metacyc))
colnames(metacyc)<-gsub("X", "",  colnames(metacyc))

# Sum strains within species
strains2 <- strains %>%
  group_by(Species) %>%
  summarise(across(.cols = 9:37, .fns = sum, na.rm = TRUE))

# Take set for graphing
bugs<-strains2[2:30]
tax<-strains2$Species

# Fix colnames
colnames(bugs)<-gsub("_.*", "", colnames(bugs))
colnames(bugs)<-gsub("X", "",  colnames(bugs))

# fix order of cols
column_order <- met$mouseID
metacyc <- metacyc[, order(match(colnames(metacyc), column_order))]
bugs <- bugs[, order(match(colnames(bugs), column_order))]
identical(colnames(bugs), as.character(met$mouseID))
identical(colnames(metacyc), as.character(met$mouseID))

bugs[is.na(bugs)] <- 0

fit1<-ldm(formula=bugs ~ Group, data=met, dist.method = "bray", seed=12345)
fit1$p.global.omni
fit1$detected.otu.omni # The OTUs significantly associated with each covariate


fit1$p.global.omni
fit1$detected.otu.omni # The OTUs significantly associated with each covariate

w1 = match(fit1$detected.otu.omni[[1]], names(fit1$q.otu.omni))
o = w1[order(fit1$p.otu.omni[w1])]
summary.tab = data.frame(raw.pvalue=signif(fit1$p.otu.omni[o],3),
                       adj.pvalue=signif(fit1$q.otu.omni[o],3),
                       mean.freq=signif(fit1$mean.freq[o],3),
                       direction=t(ifelse(fit1$beta>0, "+", "-"))[o],
                       otu.name=names(fit1$q.otu.omni)[o], # can use taxonomy assignment
                       row.names=NULL)
colnames(summary.tab)[4] = paste("direction.", rownames(fit1$beta), sep="")

# fix taxonomy
tax_index<-data.frame(tax)
tax_index<- cbind(row_number = 1:nrow(tax_index), tax_index)
tax_index$row_number = as.character(tax_index$row_number)
colnames(tax_index)[2] = "Species"
summary.tab2<-left_join(summary.tab, tax_index, by = c("otu.name"="row_number"))

write.csv(summary.tab2, "../results/Male4-Group-associated-ldm.csv")

# Now do metacyc
fit1<-ldm(formula=metacyc ~ Group, data=met, dist.method = "bray", seed=12345)
fit1$p.global.omni
fit1$detected.otu.omni # The OTUs significantly associated with each covariate


fit1$p.global.omni
fit1$detected.otu.omni # The OTUs significantly associated with each covariate

w1 = match(fit1$detected.otu.omni[[1]], names(fit1$q.otu.omni))
o = w1[order(fit1$p.otu.omni[w1])]
summary.tab = data.frame(raw.pvalue=signif(fit1$p.otu.omni[o],3),
                       adj.pvalue=signif(fit1$q.otu.omni[o],3),
                       mean.freq=signif(fit1$mean.freq[o],3),
                       direction=t(ifelse(fit1$beta>0, "+", "-"))[o],
                       otu.name=names(fit1$q.otu.omni)[o], # can use taxonomy assignment
                       row.names=NULL)
colnames(summary.tab)[4] = paste("direction.", rownames(fit1$beta), sep="")

# fix taxonomy
tax_index<-data.frame(mtax)
tax_index<- cbind(row_number = 1:nrow(tax_index), tax_index)
tax_index$row_number = as.character(tax_index$row_number)
colnames(tax_index)[2] = "Taxonomy"
summary.tab2<-left_join(summary.tab, tax_index, by = c("otu.name"="row_number"))

write.csv(summary.tab2, "../results/Male4-Group-associated-pwy-ldm.csv")
```

Generate SupFig 4 from LDM outputs while we still have proper strains2 and metadata file
```{r}
# Load LDM results for sex
data<-read.csv("../results/Male4-Group-associated-ldm.csv")
data <- arrange(data, desc(mean.freq))
data$p.adj<-ifelse(data$adj.pvalue < 0.01, "<0.01", ifelse(data$adj.pvalue < 0.05, "<0.05", "0.05 - 0.1"))

# Plot only significant species from LDM analysis
toplot<-strains2 %>% filter(Species %in% data$Species)
toplot$Taxonomy.Id = NULL

colnames(toplot)<-gsub("_.*", "", colnames(toplot))
colnames(toplot)<-gsub("X", "",  colnames(toplot))

# Reshape data
data_long <- pivot_longer(toplot, cols = -c(Species), names_to = "subject", values_to = "value")
# add metadata back for groupings

met$mouseID<-as.character(met$mouseID)
colnames(data_long)[2] = "mouseID"
df3 <- data_long %>% left_join(met, by = "mouseID") %>% select(mouseID, Group, Species, value)

# Calculate freqs & SE
freqs <- df3 %>% group_by(Species, Group) %>%
  summarize(mean=mean(value), sd=sd(value))
freqs$se = freqs$sd / sqrt(10)
freqs<-left_join(freqs, data)
freqs <- arrange(freqs, desc(mean.freq))
freqs$sig<-ifelse(freqs$adj.pvalue < 0.001, "***", ifelse(freqs$adj.pvalue < 0.01, "**", (ifelse(freqs$adj.pvalue < 0.05, "*", "-"))))

# Make plot
p<-ggplot(freqs,  aes(x = reorder(Species, mean.freq), y     = mean, color = Group)) + geom_point() + geom_errorbar(aes(ymin  = mean - se, ymax  = mean + se))  + coord_flip()  + scale_y_sqrt() + xlab("Species") + ylab("Mean Abundance") +   scale_colour_manual(values=c("forestgreen", "red", "blue"))

p + geom_text(aes(y = 0.23, x = freqs$Species), label = freqs$sig, color="black")

ggsave("../results/SupFig5.pdf", width=12, height=8)
```

Generate SupFig 5 from LDM outputs while we still have proper strains2 and metadata file
```{r}
# Load LDM results for sex
data<-read.csv("../results/Male4-Group-associated-pwy-ldm.csv")
data <- arrange(data, desc(mean.freq))
data$p.adj<-ifelse(data$adj.pvalue < 0.01, "<0.01", ifelse(data$adj.pvalue < 0.05, "<0.05", "0.05 - 0.1"))
data<-data %>% filter(adj.pvalue < 0.01)

metacyc$Taxonomy = mtax
# Plot only significant species from LDM analysis
toplot<-metacyc %>% filter(Taxonomy %in% data$Taxonomy) 

# Reshape data
data_long <- pivot_longer(toplot, cols = -c(Taxonomy), names_to = "subject", values_to = "value")
# add metadata back for groupings

met$mouseID<-as.character(met$mouseID)
colnames(data_long)[2] = "mouseID"
df3 <- data_long %>% left_join(met, by = "mouseID") %>% select(mouseID, Group, Taxonomy, value)

# Calculate freqs & SE
freqs <- df3 %>% group_by(Taxonomy, Group) %>%
  summarize(mean=mean(value), sd=sd(value))
freqs$se = freqs$sd / sqrt(10)
freqs<-left_join(freqs, data)
freqs <- arrange(freqs, desc(mean.freq))
freqs$sig<-ifelse(freqs$adj.pvalue < 0.001, "***", ifelse(freqs$adj.pvalue < 0.01, "**", (ifelse(freqs$adj.pvalue < 0.05, "*", "-"))))

# Make plot
p<-ggplot(freqs,  aes(x = reorder(Taxonomy, mean.freq), y     = mean, color = Group)) + geom_point() + geom_errorbar(aes(ymin  = mean - se, ymax  = mean + se))  + coord_flip()  + scale_y_sqrt() + xlab("Metacyc Pathway") + ylab("Mean Abundance") +   scale_colour_manual(values=c("forestgreen", "red", "blue"))

p + geom_text(aes(y = 0.23, x = freqs$Taxonomy), label = freqs$sig, color="black")

ggsave("../results/SupFig4.pdf", width=12, height=8)
```

```{r}
sessionInfo()
```

